Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-11-18 bbbq_article.tex
Processed feedback Maksim. Thanks Maksim!

\maksim{
  I would change the whole sentence: 
  T cells are an important part of the acquired immunity, they specialize in recognition of pathogen-derived peptides, called epitopes, presented on the cells surface of antigen presenting cells (APC) within the Major Histocompatibility Complexes (MHC) class I and II.
}
The MHC proteins are heterodimeric complexes encoded by the
HLA (Human Leukocyte Antigens) genes.
In humans, the peptide binding groove of MHC-I is made by only the alpha subunit. There are three classical haplotypes of MHC-I, hallmarked by a highly polymorphic alpha chain called HLA-A, HLA-B and HLA-C, that all present epitopes to cytolytic T cells. 
For MHC-II, both the alpha and the beta chains contribute to the peptide binding groove. There are three classical haplotypes of MHC-II as well, called HLA-DR, HLA-DQ and HLA-DP, that all present epitopes to helper T cells.
Each MHC complex can present a subset of all possible peptides.
For example, HLA-A and HLA-B have no overlap in which
epitopes they bind \cite{lund2004definition}.
Moreover, the HLA genes of humans are highly polymorphic, with hundreds 
to thousands of different alleles, 
and each different allele presents a different subset of peptides \cite{marsh2010nomenclature}.

@@

Many studies are aimed at identifying the repertoire of epitopes that are presented 
in any of the different alleles to determine which epitopes will result in an immune response, 
as this will for instance aid the design of vaccines. 
These studies have led to the development of prediction algorithms 
that allow for very reliable \emph{in silico} 
\maksim{predictions of the peptide binding affinities}
predictions of the binding affinities of peptides
\cite{larsen2010identification,schellens2008unanticipated,tang2011genome}.
For example, S. Tang et al. \cite{tang2011genome} found that, 
of the 432 peptides that were predicted to bind to an MHC allele,
86\% were experimentally confirmed to do so.

@@

Using these prediction algorithms, 
we recently showed that peptides derived 
from transmembrane helices (TMHs) 
are likely to be more frequently presented by MHC-I 
than expected based on their abundance \cite{bianchi2017},
which is in line with a previous study 
by Istrail et al \cite{istrail2004comparative},
demonstrating that N-terminal signal sequences 
are likely to be presented within major histocompatibility complexes, 
due their hydrophobic nature. 
Moreover, we showed that some well-known immunodominant peptides stem from TMHs. 
This over-presentation is attributed to the fact 
that the peptide-binding groove of most MHC-I alleles
is relatively hydrophobic, 
and therefore hydrophobic TMH-derived peptides have a higher affinity 
to bind than their soluble hydrophilic 
\maxim{shouldn't be hydrophobic? 
  A hydrophobic peptide binds to its hydrophobic counterpart - hyprophobic grove of MHC-I}
counterparts.

TMHs are hydrophobic 
as they need to span the hydrophobic lipid bilayer of cellular membranes.
They consist of an alpha helix of, on average, 23 amino acids in length. 
TMHs can also be predicted with high accuracy from a protein sequence 
by bioinformatics approaches \cite{krogh2001predicting,kall2004combined,arai2004conpred,jones2007improving,klammer2009metatm,wang2019efficient}. For example, a study by Jones \cite{jones2007improving} found that,
from 184 transmembrane proteins (TMPs) with known topology, 
80\% of the TMH predictions 
\maxim{this part is confusing: 
  correct maybe:
  for example, TMH prediction algorithms identified 80\% of TMH in 184 transmembrane proteins (TMPs) with known topology [12].}
replicated this finding.

TMHs are common structures in the proteins of humans and microbes. 
Different TMH prediction tools estimate that 15-39\% of all proteins 
in the human proteome contain at least one TMH \cite{ahram2006estimation}.
However, the physiological reason why peptides derived from TMHs 
would be presented more often than peptides 
stemming from soluble (i.e., extracellular or cytoplasmic) protein regions is unknown. 
In this study, we hypothesized that the presentation of 
TMH residues is evolutionarily preferred, 
since TMHs are less prone to undergo escape mutations. 
One reason to expect such a reduced 
variability (and hence evolutionary conservation) in TMHs, 
is that these are restricted in their variability 
by the functional requirement to span a lipid bilayer. 
This limits 
many of the amino acids present in TMHs 
to have hydrophobic side chains 
\cite{hessa2007molecular,jones1994model}.
Therefore, we speculated that the TMHs of pathogens 
might have a lower chance to develop escape mutations, 
as that will result in a dysfunctional TMH 
and render the protein inactive.

This study had two objectives. 
First, we aimed to generalize our findings by predicting
\maxim{the antigenic presentation}
the presentation of peptides 
from different kingdoms of life in both MHC-I and -II. 
From these \emph{in silico} predictions, we conclude that TMH-derived
epitopes from a human, viral and bacterial proteome
are likely to be presented more often than expected by chance 
for most alleles of MHC-I and II. 
We confirmed the presentation of TMH-derived peptides 
by re-analysis of peptide elution studies. 
Second, we tested our hypothesis that TMHs 
are more evolutionary conserved than soluble protein regions.
Our analysis of human single nucleotide polymorphisms (SNPs) showed 
that random point mutations are indeed less likely
to occur within TMHs. 
These findings strengthen the emerging notion 
that TMHs are important for the T cell-mediated adaptive immune system, 
and hence are of  
overlooked importance in vaccine development.

@@

For MHC-I, 9-mers were used, 
as this is the length most frequently presented in MHC-I 
and was used in our earlier study \cite{bianchi2017}. 
For MHC-II, 14-mers were used, 
as this is the most frequently occurring epitope length \cite{bergseng2015different}.
A human (UniProt ID UP000005640\_9606), 
viral (SARS-CoV-2, UniProt ID UP000464024) 
and bacterial (\emph{Mycobacterium tuberculosis}, UniProt ID UP000001584) 
reference proteome was used. 
\verb;TMHMM; \cite{krogh2001predicting} was used to predict the topology 
of the proteins within these proteomes.
To predict the affinity of an epitope to a certain HLA allele,
 \verb;EpitopePrediction; \cite{bianchi2017} for MHC-I 
and \verb;MHCnuggets; \cite{shao2020high} for MHC-II was used.
Both MCH-I and MHC-II alleles were selected to have a high occurrence
\maxim{was it meant to be: 
were selected based on their occurrence scores? or based in their prevalence in the population?},
where the alleles of the MHC-I haplotypes are the alleles representing the 13 supertypes 
with over 99.6\% coverage of the population's MHC-I repertoire as defined by \cite{lund2004definition} \cite{sette1999},
and the 21 MHC-II haplotype alleles, have a phenotypic frequency 
of 14\% or more in the human population \cite{greenbaum2011functional}.

We define a protein to be a binder if, for a certain haplotype, 
any of its 9-mer or 14-mer peptides have an IC50 value in the lowest 2\% of 
all peptides within a \emph{proteome} 
(see supplementary Tables \ref{tab:ic50_binders_mhc1} and \ref{tab:ic50_binders_mhc2} for values), 
this differs from our previous study where we defined
a binder as having an IC50 in the lowest 2\% 
of the peptides within a \emph{protein}.
% See https://github.com/richelbilderbeek/bianchi_et_al_2017/blob/72e6755a31d400158368509fd80a41e984677ab1/predict-binders.R#L17
This revised definition precludes bias of proteins 
that give rise to no or only very few MHC epitopes.
To verify that the results are similar \maxim{does it mean: that the new thresholding does not obscure the previous results, a side...}, a side by side comparison was performed shown in the supplementary materials, Figures \ref{fig:bianch_et_al_2017_1a}
and \ref{fig:bilderbeek_et_al_2021_1a}.

@@

To obtain experimental confirmation that peptides stemming from TMHs 
are presented \emph{in vivo} in MHC-I and MHC-II,
we mined the Immune Epitope Database (IEDB) \cite{vita2019immune}
for confirmed human MHC-ligands.
We queried the IEDB for all linear epitopes obtained
from MHC ligand assays in healthy humans, 
carrying the MHC alleles as used in this study.
From these epitopes, we kept those that were present
exactly once in the human reference proteome
with UniProt ID UP000005640\_9606.
We predicted the topology of the protein each epitope
was found in, using \verb;TMHMM; \cite{krogh2001predicting},
from which we concluded if the epitope is overlapping with a TMH
with at least 1 amino acid \maxim{1 amino acid overlap is already a good enough evidence to label something as TMH?}.

@@

To determine the evolutionary conservation of TMHs,
human single nucleotide polymorphisms (SNPs) were first collected
\maxim{..TMHs, we first collected (SNPs) resulting in a single amino acid substitution to determine if this occurred within a predicted TMH or not.}
that resulted in a single amino acid substitution,
and we then determined if this substitution occurred within a predicted TMH or not.

@@

The number of SNPs was limited to the first 250 variations per gene,
resulting in $\approx$61k variations.
Only variations that result in a SNP for
a single amino acid substitution were analyzed, resulting in $\approx$38k SNPs.
The exact amounts can be found in the supplementary materials,
Tables \maxim{earlier you used brackets to refer to these tables} \ref{tab:ncbi_counts_1} and \ref{tab:ncbi_counts_2}.

@@

Figure \ref{fig:bbbq_1_smart_results} shows the predicted presentation of TMH-derived peptides in MHC-I,
for a human, viral and bacterial proteome.
Per MHC-I allele, it shows the percentage of binders that overlap with a TMH 
with at least one residue.
The horizontal line shows the expected percentage of TMH-derived epitopes 
that would be presented, if TMH-derived epitopes would be presented just as 
likely as epitopes derived from soluble regions \maxim{assuming equal incidence of soluble and TMH-derived epitope presentation.}.
For 11 out of 13 MHC-I alleles, TMH-derived epitopes are predicted to be presented more often 
than the null expectation, for a human and bacterial proteome.
For the viral proteome, 12 out of 13 MHC-I alleles present
TMH-derived epitopes more often than expected by chance.
The extent of the over-presentation between the different alleles
is similar for the probed proteomes, 
which strengthens our previous conclusion \cite{bianchi2017} 
that the hydrophobicity of the MHC-binding groove 
is the main factor responsible for the predicted over-presentation 
of TMH-derived peptides.

@@

Per protein, we calculated two percentages: 
(1) the percentage of the total protein predicted to be TMHs \maxim{is it percentage of a protein sequence length bearing TMHs? 
if it is true - it would be a better wording}, 
and (2) the percentage of SNPs located within these predicted TMHs.
Each percentage pair was plotted in figure \ref{fig:f_snps_found_and_expected}.
The proportion of SNPs found in TMHs varied from 
none (i.e., all SNPs were in
soluble regions) to all (i.e., all SNPs were in
TMHs).
To determine if SNPs were randomly distributed over the protein, we performed a linear regression analysis,
and added a 95\% confidence interval on this regression.
This linear fit nearly goes through the origin and has a slope
below the line of equality,
which shows that less SNPs are found in TMHs than expected by chance.

We determined the probability to find the observed amount
of SNPs in TMHs by chance, i.e., when assuming SNPs occur 
just as likely in soluble domains as in TMHs.
We used a binomial Poisson distribution, 
where the number of trails ($n$) equals the number of SNPs, 
which is 21,208. 
The probability of success for the $i$th TMP ($p\_i$), 
is the percentage of residues within a TMH per TMP. 
These percentages are shown as a histogram 
in figure \ref{fig:f_tmh_ncbi}. 
The expected number of SNPs expected to be found in 
TMHs by chance equals $\sum{p} \approx 4,141$.
As we observed 3,803 SNPs in TMHs, 
we calculated the probability of having that amount or less successes.
We used the type I error cut-off value of $\alpha = 2.5\%$.
The chance to find, within TMHs, this amount or less SNPs 
equals $6.8208 \cdot 10^{-11}$.
We determined the relevance of this finding, by
calculating how much less SNPs are found in TMHs,
when compared to soluble regions, which is the
ratio between the number of SNPs found in TMHs
versus the number of SNPs as expected by chance.
In effect, per 1000 SNPs found in soluble protein domains, 
one finds 918 SNPs in TMHs,
\maxim{your current figure does not describe it}
as depicted in figure \ref{fig:conservation}.

>>>>

The MHC proteins are heterodimeric complexes encoded by the
HLA (Human Leukocyte Antigens) genes.
In humans, the peptide binding groove of MHC-I is made by only the alpha subunit. There are three classical haplotypes of MHC-I, hallmarked by a highly polymorphic alpha chain called HLA-A, HLA-B and HLA-C, that all present epitopes to cytolytic T cells. 
For MHC-II, both the alpha and the beta chains contribute to the peptide binding groove. There are three classical haplotypes of MHC-II as well, called HLA-DR, HLA-DQ and HLA-DP, that all present epitopes to helper T cells.
Each MHC complex can present a subset of all possible peptides.
For example, HLA-A and HLA-B have no overlap in which
epitopes they bind \cite{lund2004definition}.
Moreover, the HLA genes of humans are highly polymorphic, with hundreds 
to thousands of different alleles, 
and each different allele presents a different subset of peptides \cite{marsh2010nomenclature}.

@@

Many studies are aimed at identifying the repertoire of epitopes that are presented 
in any of the different alleles to determine which epitopes will result in an immune response, 
as this will for instance aid the design of vaccines. 
These studies have led to the development of prediction algorithms 
that allow for very reliable \emph{in silico} 
predictions of the peptide binding affinities
\cite{larsen2010identification,schellens2008unanticipated,tang2011genome}.
For example, S. Tang et al. \cite{tang2011genome} found that, 
of the 432 peptides that were predicted to bind to an MHC allele,
86\% were experimentally confirmed to do so.

@@

Using these prediction algorithms, 
we recently showed that peptides derived 
from transmembrane helices (TMHs) 
are likely to be more frequently presented by MHC-I 
than expected based on their abundance \cite{bianchi2017},
which is in line with a previous study 
by Istrail et al \cite{istrail2004comparative},
demonstrating that N-terminal signal sequences 
are likely to be presented within major histocompatibility complexes, 
due their hydrophobic nature. 
Moreover, we showed that some well-known immunodominant peptides stem from TMHs. 
This over-presentation is attributed to the fact 
that the peptide-binding groove of most MHC-I alleles
is relatively hydrophobic, 
and therefore hydrophobic TMH-derived peptides have a higher affinity 
to bind than their soluble hydrophobic 
counterparts.

TMHs are hydrophobic 
as they need to span the hydrophobic lipid bilayer of cellular membranes.
They consist of an alpha helix of, on average, 23 amino acids in length. 
TMHs can also be predicted with high accuracy from a protein sequence 
by bioinformatics approaches \cite{krogh2001predicting,kall2004combined,arai2004conpred,jones2007improving,klammer2009metatm,wang2019efficient}. 
For example, a study by Jones \cite{jones2007improving} found that,
from 184 transmembrane proteins (TMPs) with known \emph{in vitro} topology, 
80\% of the TMH predictions of these proteins
matched the experimental findings.
TMHs are common structures in the proteins of humans and microbes. 
Different TMH prediction tools estimate that 15-39\% of all proteins 
in the human proteome contain at least one TMH \cite{ahram2006estimation}.
However, the physiological reason why peptides derived from TMHs 
would be presented more often than peptides 
stemming from soluble (i.e., extracellular or cytoplasmic) protein regions is unknown. 
In this study, we hypothesized that the presentation of 
TMH residues is evolutionarily preferred, 
since TMHs are less prone to undergo escape mutations. 
One reason to expect such a reduced 
variability (and hence evolutionary conservation) in TMHs, 
is that these are restricted in their variability 
by the functional requirement to span a lipid bilayer. 
This limits 
many of the amino acids present in TMHs 
to have hydrophobic side chains 
\cite{hessa2007molecular,jones1994model}.
Therefore, we speculated that the TMHs of pathogens 
might have a lower chance to develop escape mutations, 
as that will result in a dysfunctional TMH 
and render the protein inactive.

This study had two objectives. 
First, we aimed to generalize our findings by predicting
the antigenic presentation
from different kingdoms of life in both MHC-I and -II. 
From these \emph{in silico} predictions, we conclude that TMH-derived
epitopes from a human, viral and bacterial proteome
are likely to be presented more often than expected by chance 
for most alleles of MHC-I and II. 
We confirmed the presentation of TMH-derived peptides 
by re-analysis of peptide elution studies. 
Second, we tested our hypothesis that TMHs 
are more evolutionary conserved than soluble protein regions.
Our analysis of human single nucleotide polymorphisms (SNPs) showed 
that random point mutations are indeed less likely
to occur within TMHs. 
These findings strengthen the emerging notion 
that TMHs are important for the T cell-mediated adaptive immune system, 
and hence are of  
overlooked importance in vaccine development.

@@

For MHC-I, 9-mers were used, 
as this is the length most frequently presented in MHC-I 
and was used in our earlier study \cite{bianchi2017}. 
For MHC-II, 14-mers were used, 
as this is the most frequently occurring epitope length \cite{bergseng2015different}.
A human (UniProt ID UP000005640\_9606), 
viral (SARS-CoV-2, UniProt ID UP000464024) 
and bacterial (\emph{Mycobacterium tuberculosis}, UniProt ID UP000001584) 
reference proteome was used. 
\verb;TMHMM; \cite{krogh2001predicting} was used to predict the topology 
of the proteins within these proteomes.
To predict the affinity of an epitope to a certain HLA allele,
 \verb;EpitopePrediction; \cite{bianchi2017} for MHC-I 
and \verb;MHCnuggets; \cite{shao2020high} for MHC-II was used.
Both MCH-I and MHC-II alleles were selected 
to have a high prevalence in the population,
where the alleles of the MHC-I haplotypes are the alleles representing the 13 supertypes 
with over 99.6\% coverage of the population's MHC-I repertoire as defined by \cite{lund2004definition} \cite{sette1999},
and the 21 MHC-II haplotype alleles, have a phenotypic frequency 
of 14\% or more in the human population \cite{greenbaum2011functional}.

We define a protein to be a binder if, for a certain haplotype, 
any of its 9-mer or 14-mer peptides have an IC50 value in the lowest 2\% of 
all peptides within a \emph{proteome} 
(see supplementary Tables \ref{tab:ic50_binders_mhc1} and \ref{tab:ic50_binders_mhc2} for values), 
this differs from our previous study where we defined
a binder as having an IC50 in the lowest 2\% 
of the peptides within a \emph{protein}.
% See https://github.com/richelbilderbeek/bianchi_et_al_2017/blob/72e6755a31d400158368509fd80a41e984677ab1/predict-binders.R#L17
This revised definition precludes bias of proteins 
that give rise to no or only very few MHC epitopes.
To verify that the slight change in method yields similar results,
a side by side comparison is shown in the supplementary materials, 
Figures \ref{fig:bianch_et_al_2017_1a} and \ref{fig:bilderbeek_et_al_2021_1a}.

@@

To obtain experimental confirmation that peptides stemming from TMHs 
are presented \emph{in vivo} in MHC-I and MHC-II,
we mined the Immune Epitope Database (IEDB) \cite{vita2019immune}
for confirmed human MHC-ligands.
We queried the IEDB for all linear epitopes obtained
from MHC ligand assays in healthy humans, 
carrying the MHC alleles as used in this study.
From these epitopes, we kept those that were present
exactly once in the human reference proteome
with UniProt ID UP000005640\_9606.
We predicted the topology of the protein each epitope
was found in, using \verb;TMHMM; \cite{krogh2001predicting},
from which we concluded if the epitope is overlapping with a TMH
with at least 1 amino acid.

@@

To determine the evolutionary conservation of TMHs,
we first collected human single nucleotide polymorphisms (SNPs) 
resulting in a single amino acid substitution to determine if this occurred within a predicted TMH or not.

@@

The number of SNPs was limited to the first 250 variations per gene,
resulting in $\approx$61k variations.
Only variations that result in a SNP for
a single amino acid substitution were analyzed, resulting in $\approx$38k SNPs.
The exact amounts can be found in the supplementary materials,
Tables \ref{tab:ncbi_counts_1} and \ref{tab:ncbi_counts_2}.

@@

Figure \ref{fig:bbbq_1_smart_results} shows the predicted presentation of TMH-derived peptides in MHC-I,
for a human, viral and bacterial proteome.
Per MHC-I allele, it shows the percentage of binders that overlap with a TMH 
with at least one residue.
The horizontal line shows the expected percentage of TMH-derived epitopes 
that would be presented, if TMH-derived epitopes would be presented just as 
likely as epitopes derived from soluble regions,
when assuming equal incidence of soluble and TMH-derived epitope presentation.
For 11 out of 13 MHC-I alleles, TMH-derived epitopes are predicted to be presented more often 
than the null expectation, for a human and bacterial proteome.
For the viral proteome, 12 out of 13 MHC-I alleles present
TMH-derived epitopes more often than expected by chance.
The extent of the over-presentation between the different alleles
is similar for the probed proteomes, 
which strengthens our previous conclusion \cite{bianchi2017} 
that the hydrophobicity of the MHC-binding groove 
is the main factor responsible for the predicted over-presentation 
of TMH-derived peptides.

@@

Per protein, we calculated two percentages: 
(1) the percentage of a protein sequence length bearing TMHs, 
and (2) the percentage of SNPs located within these predicted TMHs.
Each percentage pair was plotted in figure \ref{fig:f_snps_found_and_expected}.
The proportion of SNPs found in TMHs varied from 
none (i.e., all SNPs were in
soluble regions) to all (i.e., all SNPs were in
TMHs).
To determine if SNPs were randomly distributed over the protein, we performed a linear regression analysis,
and added a 95\% confidence interval on this regression.
This linear fit nearly goes through the origin and has a slope
below the line of equality,
which shows that less SNPs are found in TMHs than expected by chance.

We determined the probability to find the observed amount
of SNPs in TMHs by chance, i.e., when assuming SNPs occur 
just as likely in soluble domains as in TMHs.
We used a binomial Poisson distribution, 
where the number of trails ($n$) equals the number of SNPs, 
which is 21,208. 
The probability of success for the $i$th TMP ($p\_i$), 
is the percentage of residues within a TMH per TMP. 
These percentages are shown as a histogram 
in figure \ref{fig:f_tmh_ncbi}. 
The expected number of SNPs expected to be found in 
TMHs by chance equals $\sum{p} \approx 4,141$.
As we observed 3,803 SNPs in TMHs, 
we calculated the probability of having that amount or less successes.
We used the type I error cut-off value of $\alpha = 2.5\%$.
The chance to find, within TMHs, this amount or less SNPs 
equals $6.8208 \cdot 10^{-11}$.
We determined the relevance of this finding, by
calculating how much less SNPs are found in TMHs,
when compared to soluble regions, which is the
ratio between the number of SNPs found in TMHs
versus the number of SNPs as expected by chance.
In effect, per 1000 SNPs found in soluble protein domains, 
one finds 918 SNPs in TMHs,
as depicted (as percentages) in figure \ref{fig:conservation}.
